### 2015-07-30 manuscript/Revision/Text_S1.tex
Commit prior to revision of PLoS One ms. Switched Bayesian GAM out for frequentist one using mgcv, changed color scheme for plot_N_map

We write all statistical models for ecological prediction in the form
\begin{linenomath*}
\begin{equation}
  \label{eq:basic_model}
  Y_i \sim f_Y(g^{-1}(\mu_i)),
\end{equation}
\end{linenomath*}
where $\mu_i=\xi_i + \epsilon_i$, $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function,
$\xi_i$ is a linear predictor, and $\epsilon_i \sim {\rm Normal}(0,\tau_\epsilon)$ is iid Gaussian error with precision parameter $\tau_\epsilon$.  The specification in Eq. \ref{eq:basic_model} is thus doubly stochastic, in the sense that we assume error associated with $f_Y$ as well as in the location parameter $\mu_i$.  This setup can be useful computationally, and can also be used to approximate singly stochastic system by setting $\tau_\epsilon$ to a large value.  For instance, models for count data often assume a Poisson error structure with a log link function to ensure that the Poisson intensity parameter is greater than zero.  In this case, we would specify
\begin{linenomath*}
\begin{equation*}
  Y_i \sim \mathrm{Poisson}(\exp(\xi_i + \epsilon_i)),
\end{equation*}
\end{linenomath*}
a configuration known as a log-Gaussian Cox process.
We now describe how different classes of statistical models can be developed depending on how one structures the linear predictor, $\xi_i$.

@@

\hspace{.5in}There are a number of ways smooth relationships can be modeled; in order to formulate a GAM in the notation of Eq. \ref{eq:basic_model} we employ a knot-based kernel smoother with a Gaussian radial basis function.  The basic notion is to place $k_j$ knots, which we will denote by $\boldsymbol{\omega}_j = \{ \omega_{1j},\omega_{2j},\hdots,\omega_{k_j j} \}$, throughout the range of a given regressor $j$, where the number of knots controls the level of smoothing.  For each regressor of interest, we can then calculate a matrix ${\bf K}_j$ of dimension $(n,k_j)$ (where $n$ gives the number of observed data locations), with entries $z_{ik}=\mathcal{N}(X_{ij};\omega_{kj},\tau_j)$, where $\mathcal{N}(x;\mu,\tau)$ denotes a Gaussian distribution with mean $\mu$ and precision $\tau$, $i$ indexes a particular observed data location, and $k$ denotes the $k$th knot for covariate $j$, $k \in \{1,2,\hdots,k_j \}$. We can then introduce additional regression coefficients $\boldsymbol{\alpha}_j$ to help model the smooth relationship, and incorporate these into the linear predictor.  In the case of one smooth term, we then have  in Eq. \ref{eq:basic_model}:
\begin{linenomath*}
\begin{equation}
  \label{eq:GAM}
  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + {\bf K}_j \boldsymbol{\alpha}_j
\end{equation}
\end{linenomath*}
When fitting such a model, one needs to select the number and location of knots, as well as the precision parameter $\tau_{\omega,j}$ for the Gaussian basis kernels.  This could be done using some notion of optimality or prediction error (as with cross validation).  In practice, a simple recommendation is to set $\tau_{\omega,j}$ less than the range of the modeled covariate, and greater than the typical difference between observed values of the covariate.

\hspace{.5in}Our main point in writing the GAM as in Eq. \ref{eq:GAM} is to emphasize the structural similarity between GLMs and certain classes
of GAMs.  In particular,
we can rewrite Eq. \ref{eq:GAM} as
$\boldsymbol{\xi}={\bf X}_{aug} \boldsymbol{\beta}_{aug}$,
where ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K}_j \right]$ (i.e., concatenating ${\bf X}$ and ${\bf K}_j$ horizontally), and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha}_j \right] ^\prime$ (the subscript $aug$ denotes augmentation).  We shall exploit this structure when examining
extrapolation bias.  We note that similar formulations for GAMs (e.g. using thin plate splines) can be represented in a similar form \citep[cf.][]{CrainiceanuEtAl2005}.

@@

The previous two modeling frameworks (GLMs and GAMs) do not acknowledge spatial autocorrelation above and beyond that induced
by modeled covariates.  However, it is common for residuals from GLM and GAM model fits to include spatial autocorrelation, which violates their common assumption of independently distributed error \citep{Legendre1993,LichsteinEtAl2002}.  Parameter estimates from GLMs and GAMs that display residual autocorrelation should be interpreted with caution, as they will tend to have overstated precision and may even be biased.  In such situations, analysts often employ spatial regression models which explicitly account for spatial autocorrelation above and beyond that explained by modeled covariates.

@@

We now describe methods for conducting Bayesian analysis for the models described above.  We utilize Gibbs sampling, characterized by cyclically sampling groups of parameters according to their so-called full conditional distributions conditionally on data and other parameters \citep{GelmanEtAl2004}.  In some cases, we are able to solve for full conditional distributions analytically and to sample from known families of probability distributions; in others, we resort to Metropolis-Hastings steps.

@@

\subsection{Calculating the gIVH}

In order to calculate prediction variance on the real scale for generalized models, one can use a procedure such as the delta method \citep{Dorfman1938}, or in the case of Bayesian inference, can simply calculate posterior predictive variance of $\boldsymbol{\lambda}_x$.  For the delta method approach, we can write the variance of the Poisson expectation as
\begin{linenomath*}
\begin{eqnarray}
  \label{eq:var-lambda}
  {\rm var}(\boldsymbol{\lambda}_x) & = & {\rm var}(g(\boldsymbol{\mu}_x)) \\
  & \approx & \boldsymbol{\Delta} {\rm var}(\boldsymbol{\mu}_x)\boldsymbol{\Delta}^\prime,
%  \boldsymbol{\Delta} & = & \left[ \begin{array}{cccc}
%      \frac{\partial g(\mu_1)}{\partial \mu_1} & \frac{\partial g(\mu_1)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_1)}{\partial \mu_n} \\
%      \frac{\partial g(\mu_2)}{\partial \mu_1} & \frac{\partial g(\mu_2)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_2)}{\partial \mu_n} \\
%      \vdots & \vdots & \ddots & \vdots \\
%      \frac{\partial g(\mu_n)}{\partial \mu_1} & \frac{\partial g(\mu_n)}{\partial \mu_2} & %\hdots & \frac{\partial g(\mu_n)}{\partial \mu_n} \\
%  \end{array} \right].
\end{eqnarray}
\end{linenomath*}
where $\boldsymbol{\Delta}$ is a matrix of partial derivatives $\partial g(\mu_r)/\partial \mu_c$ ($r$ and $c$ denoting rows and columns of $\boldsymbol{\Delta}$, respectively).  Under common univariate link functions (e.g. log, logit, probit), $\boldsymbol{\Delta}$ has a diagonal form, while for multivariate links (e.g. multinomial logit) $\boldsymbol{\Delta}$ will be dense.

\hspace{.5in} Unlike Cook's original formulation, the gIVH will often involve unknown parameters.  Further, the exact form of ${\rm var}(\hat{\boldsymbol{\beta}}_{aug})$ depends on the underlying model structure and estimation procedure.  We impose a probability distribution $[\boldsymbol{\theta}]$ for unknown parameters, and compute prediction variance as
\begin{linenomath*}
\begin{equation}
  \label{eq:E_var_lambda}
  {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}) = \int_{\boldsymbol{\theta}} {\rm var}(\boldsymbol{\lambda}_x | {\bf Y}, \boldsymbol{\theta}) [\boldsymbol{\theta}] d \boldsymbol{\theta}.
\end{equation}
\end{linenomath*}

\hspace{.5in} Our approach in this paper is to conduct a Bayesian analysis (using the Gibbs sampling algorithms outlined above) and to use posterior predictive inference to summarize ${\rm var}(\boldsymbol{\lambda}_x | {\bf Y})$.  Although we use a Bayesian formulation for modeling convenience, one could also consider substituting parametric approximations for ${\rm Var}(\mu_x)$ and $\mu_x$ based on output from standard likelihood-based analyses.

>>>>

In this paper, we restrict consideration to statistical models for ecological prediction to those that can be written with the form
\begin{linenomath*}
\begin{equation}
  \label{eq:basic_model}
  Y_i \sim f_Y(g^{-1}(\mu_i)),
\end{equation}
\end{linenomath*}
where $\mu_i=\xi_i + \epsilon_i$, $f_Y$ denotes a probability density or mass function (e.g. Bernoulli, Poisson), $g$ gives a link function, and
$\xi_i$ is a linear predictor.  Many models can be fit with $\epsilon_i = 0$ (such as many frequentist generalized linear and generalized additive models), but the Bayesian models that we develop subsequently involve the specification $\epsilon_i \sim {\rm Normal}(0,\tau_\epsilon)$ where $\tau_\epsilon$ is a precision parameter.  This specification is doubly stochastic, in the sense that we assume error associated with $f_Y$ as well as in the location parameter $\mu_i$.  This setup can be useful computationally, and can also be used to approximate singly stochastic system by setting $\tau_\epsilon$ to a large value.  For instance, models for count data often assume a Poisson error structure with a log link function to ensure that the Poisson intensity parameter is greater than zero.  In this case, we would specify
\begin{linenomath*}
\begin{equation*}
  Y_i \sim \mathrm{Poisson}(\exp(\xi_i + \epsilon_i)),
\end{equation*}
\end{linenomath*}
a configuration known as a log-Gaussian Cox process.
We now describe how different classes of statistical models can be developed depending on how one structures the linear predictor, $\xi_i$.

@@

\hspace{.5in}There are a number of ways smooth relationships can be modeled, depending upon the type of basis function employed (e.g. cubic smoothing or thin plate splines). Regardless of basis choice, it is often possible to write GAMs by substituting the following linear predictor into Eq. \ref{eq:basic_model}:
\begin{linenomath*}
\begin{equation}
  \label{eq:GAM}
  \boldsymbol{\xi}={\bf X} \boldsymbol{\beta} + {\bf K} \boldsymbol{\alpha} \end{equation}
\end{linenomath*}
Here, the component ${\bf X} \boldsymbol{\beta}$ correspond to fixed effects in the usual GLM sense (which may or may not be modeled), ${\bf K}$ represents a smooth basis matrix, and $\boldsymbol{\alpha}$ is a vector of additional parameters.  Our main point in writing the GAM as in Eq. \ref{eq:GAM} is to emphasize the structural similarity between GLMs and certain classes
of GAMs.  In particular,
we can rewrite Eq. \ref{eq:GAM} as
\begin{linenomath*}
  \begin{equation}
    \label{eq:GAM-aug}
    \boldsymbol{\xi}={\bf X}_{aug} \boldsymbol{\beta}_{aug},
  \end{equation}
\end{linenomath*}
where ${\bf X}_{aug} = \left[ {\bf X} \hspace{3mm} {\bf K} \right]$ (i.e., concatenating ${\bf X}$ and ${\bf K}$ horizontally), and $\boldsymbol{\beta}_{aug}=\left[  \boldsymbol{\beta} \hspace{3mm} \boldsymbol{\alpha} \right] ^\prime$ (the subscript $aug$ denotes augmentation).

@@

The previous two modeling frameworks (GLMs and GAMs) do not acknowledge spatial autocorrelation above and beyond that induced
by modeled covariates.  However, it is common for residuals from GLM and GAM model fits to include spatial autocorrelation, which violates their common assumption of independently distributed error \citep{Legendre1993,LichsteinEtAl2002}.  GLMs and GAMs that display residual autocorrelation should be interpreted with caution, as they will tend to have overstated precision and may even be biased.  In such situations, analysts often employ spatial regression models which explicitly account for spatial autocorrelation above and beyond that explained by modeled covariates.

@@

We now describe methods for conducting Bayesian analysis for the GLM and STRM models described above (note that we used the \texttt{mgcv} R package of \citet{Wood2006} to conduct frequentist analysis of the GAM model, as described in the next section).  We utilize Gibbs sampling, characterized by cyclically sampling groups of parameters according to their so-called full conditional distributions conditionally on data and other parameters \citep{Gelman2004}.  In some cases, we are able to solve for full conditional distributions analytically and to sample from known families of probability distributions; in others, we resort to Metropolis-Hastings steps.

@@

\section{Frequentist analysis using \texttt{mgcv} package}

In this section, we describe analysis using the \texttt{mgcv} R package \citep{Wood2006}, and how the gIVH can be calculated conditioned on a fitted GAM model.  In particular, we calculate that gIVH using prediction variance with the following steps:
\begin{enumerate}
  \item Fit a GAM model using the \texttt{gam} function.  For this  presentation, we will assume this object is given the name ``gam.fit."  If the proportion of each sample unit (e.g., grid cell) that is surveyed is less than 1.0, an offset should be included in the formula object required by the \texttt{gam} function.  For instance, a count model might specify a formula object such as \texttt{count \textasciitilde offset(log(Offset)) + s(my.cov,5)}.  In this case \texttt{Offset} is a vector holding the proportion of each sample unit that is surveyed.
  \item Extract the augmented design matrix ${\bf X}_{aug}$ from Eq. \ref{eq:GAM-aug} for observed data using the \texttt{predict.gam} function from \texttt{mgcv}.  \citet{Wood2006} calls this the ``prediction matrix," and one can obtain it using the \texttt{type="lpmatrix"} option in \texttt{predict.gam}.
  \item Calculate variance on the linear predictor scale as $V_\mu = {\bf X}_{aug} \hat{\textrm{var}}(\hat{\boldsymbol{\beta}}_{aug}){\bf X}_{aug}^\prime$, where $\hat{\textrm{var}}(\hat{\boldsymbol{\beta}}_{aug})$ is the variance-covariance matrix for estimated parameters (obtainable using \texttt{vcov(gam.fit)}).
  \item Use the delta method \citep{Dorfman1938} to calculate the variance-covariance matrix on the response scale (call this $V_{obs}$).  Set $v_{max}$ equal to the maximum diagonal element of $V_{obs}$.
  \item Assemble a new \texttt{data.frame} (call this \texttt{Pred}), which includes covariate values for all sample units (grid cells) one wants to make inference to.
  \item Perform steps 2-3 from above using this new dataset (using \texttt{newdata=Pred} when calling predict.gam).
  \item Use the delta method \citep{Dorfman1938} to calculate the variance-covariance on the response scale for all prediction locations (call this $V_{pred}$).
  \item All sample units $i$ for which $V_{pred}[i,i] \le v_{max}$ are included in the gIVH.
\end{enumerate}

### 2015-07-30 manuscript/Revision/Text_S2.tex
Commit prior to revision of PLoS One ms. Switched Bayesian GAM out for frequentist one using mgcv, changed color scheme for plot_N_map

Spatially explicit statistical models are increasingly used to estimate animal abundance and predict species distributions from count data.  We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where predictions of animal abundance using these models may be problematic, and to determine whether statistical inferences are more robust when restricted to sampling units within the gIVH.  Each simulation replicate consisted of several steps, including
\begin{enumerate}
  \item Simulate three hypothetical, statistically dependent, spatially autocorrelated environmental covariates over a 30$\times$30 grid,
  \item Simulate animal abundance across the landscape as a function of environmental covariates,
  \item Simulate animal population surveys across the landscape, including the position of count quadrats and the resulting animal counts,
  \item Estimate animal abundance as a function of two of the environmental covariates according three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).  For the latter, only a spatial dimension was modeled.  All such models were specified hierarchically, with MCMC used for posterior simulation.
  \item Calculate the gIVH using realized posterior variance (i.e. after data were collected and analyzed).
\end{enumerate}
We now describe each of these tasks and results of the simulation study in further detail before describing results.  All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the R package \texttt{SpatPred} that accompanies this article.

@@

For each set of count data, we attempted to estimate animal abundance over the landscape using three different hierarchical models, corresponding to a generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a generalized additive model (GAM; smooth effects of covariates on the log scale), and a spatio-temporal regression model (with both linear, fixed effects and spatially autocorrelated random effects).  Further details on these models, together with the procedure used for posterior simulation, are presented in Text S1.  For this study, we used the RSR implementation of the STRM outlined in Text S1.  Prior distributions for each precision parameter, $\tau$, were set to $\text{Gamma}(1.0,0.01)$, which is diffuse while maintaining a flat shape near the origin.  Regression parameters ($\beta$) were given vague $\mathcal{N}(0,\tau=0.01)$ priors.

@@

As suggested in Text S1, we used Eq. S1.7 to calculate the gIVH, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.

>>>>

Spatially explicit statistical models are increasingly used to estimate animal abundance and predict species distributions from count data.  We employed a simulation study to examine the ability of the generalized independent variable hull (gIVH) to diagnose potential areas across the landscape where predictions of animal abundance using these models may be problematic, and to determine whether statistical inferences are more robust when restricted to sampling units within the gIVH.  Each simulation replicate consisted of several steps, including
\begin{enumerate}
  \item Simulate three hypothetical, statistically dependent, spatially autocorrelated environmental covariates over a 30$\times$30 grid,
  \item Simulate animal abundance across the landscape as a function of environmental covariates,
  \item Simulate surveys across the landscape, including the position of count quadrats and the resulting animal counts,
  \item Estimate animal abundance as a function of two of the environmental covariates according three different models: a generalized linear model (GLM), a generalized additive model (GAM), and a spatio-temporal regression model (STRM).  For the latter, only a spatial dimension was modeled.  The GLM and STRM models were specified hierarchically, with MCMC used for posterior simulation.  For the GAM, we conducted a frequentist analysis using the \texttt{mgcv} R package \citep{Wood2006}.
  \item Calculate the gIVH using realized posterior variance (i.e. after data were collected and analyzed).
\end{enumerate}
We now describe each of these tasks and results of the simulation study in further detail before describing results.  All analyses were performed in the R programming environment \citep{RTeam2012}; requisite code to recreate analyses is available in the R package \texttt{SpatPred} that accompanies this article.

@@

For each set of count data, we attempted to estimate animal abundance over the landscape using three different models, corresponding to a hierarchical, Bayesian generalized linear model (GLM; linear, fixed effects of covariates on the log scale), a frequentist generalized additive model (GAM; smooth effects of covariates on the log scale) fit using \texttt{mgcv}, and a hierarchical spatial regression model (STRM; with both linear, fixed effects and spatially autocorrelated random effects).  Further details on these models, together with the procedure used for posterior simulation, are presented in Text S1.  For this study, we used the RSR implementation of the STRM outlined in Text S1.  Prior distributions for each precision parameter, $\tau$, were set to $\text{Gamma}(1.0,0.01)$, which is diffuse while maintaining a flat shape near the origin.  Regression parameters ($\beta$) were given vague $\mathcal{N}(0,\tau=0.01)$ priors.
For the GLM and STRM, we used the median posterior prediction of total abundance as a point estimator.  For the GAM, we used the sum of predictions of total abundance on the response scale, obtained using the \texttt{predict.gam} function in \texttt{mgcv}.

@@

As suggested in Text S1, we used Eq. S1.7 to calculate the gIVH for the GLM and STRM models, substituting in samples from the joint posterior distribution of each model for $\boldsymbol{\theta}$.  For the GAM, we used the delta method \citep{Dorfman1938} to calculate prediction variance on the response scale, conditioning on estimated $\beta$ values.
